Neighborhoods and Felony Disenfranchisement: The Case of New York City

### 2020-03-04 markdown/part1/for_brennan.tex
write

\documentclass[12pt,]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
\else % if luatex or xelatex
  \ifxetex
    \usepackage{mathspec}
  \else
    \usepackage{fontspec}
  \fi
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\hypersetup{unicode=true,
            pdftitle={Neighborhoods and Felony Disenfranchisement: The Case of New York City},
            pdfauthor={Kevin Morris},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{longtable,booktabs}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

%%% Use protect on footnotes to avoid problems with footnotes in titles
\let\rmarkdownfootnote\footnote%
\def\footnote{\protect\rmarkdownfootnote}

%%% Change title format to be more compact
\usepackage{titling}

% Create subtitle command for use in maketitle
\providecommand{\subtitle}[1]{
  \posttitle{
    \begin{center}\large#1\end{center}
    }
}

\setlength{\droptitle}{-2em}

  \title{Neighborhoods and Felony Disenfranchisement: The Case of New York City}
    \pretitle{\vspace{\droptitle}\centering\huge}
  \posttitle{\par}
    \author{Kevin Morris\footnote{The author thanks Jacob Faber, Jeff Manza, Myrna Pérez, Ariel White, and Peter Miller for their comments on this project. All errors are my responsibility.}}
    \preauthor{\centering\large\emph}
  \postauthor{\par}
      \predate{\centering\large\emph}
  \postdate{\par}
    \date{November 12, 2019}

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}

\usepackage{rotating}
\newcommand{\beginsupplement}{\setcounter{table}{0}  \renewcommand{\thetable}{A\arabic{table}} \setcounter{figure}{0} \renewcommand{\thefigure}{A\arabic{figure}}}

@@

The political history of the United States has been characterized by a general, if nonlinear, trend toward universal suffrage (see, for instance, Keyssar \protect\hyperlink{ref-Keyssar2009}{2009}). At the time of the nation's founding, access to the ballot box was restricted to landed White men; over the following two centuries, the franchise was greatly expanded. Today, voting rights are considered foundational aspects of full citizenship (United Nations General Assembly Resolution 2200 (XXI)). Despite the United State's march toward ever-more-inclusive systems of democracy, however, one large group of American citizens is formally barred from voting. In most of the United States, citizens convicted of felonies are at least temporarily prohibited from casting ballots in elections (Brennan Center for Justice \protect\hyperlink{ref-bcj_laws}{2018}). Although some states such as Florida and Louisiana have gradually moved to dismantle their systems of felony disenfranchisement, an estimated 4.7 million American citizens remain disenfranchised (Uggen, Larson, and Shannon \protect\hyperlink{ref-sentencing_2016}{2016}).

Due to economic and racial segregation, these effects are highly spatially concentrated. Data available from New York City shows that in 2017, 10 of the New York Police Department's 77 precincts were responsible for more than a quarter of all arrests for felony charges. Many scholars have detailed the impact of living in areas with high levels of police activity. Residents of such neighborhoods suffer from worse physical health (Sewell and Jefferson \protect\hyperlink{ref-Sewell2016}{2016}) and are more likely to suffer from anxiety and exhibit symptoms of trauma (Geller et al. \protect\hyperlink{ref-Geller2014}{2014}). The labor markets and social networks in neighborhoods with high levels of policing and incarceration are disrupted (Clear \protect\hyperlink{ref-Clear2008}{2008}), while concentrated policing has also been credited with having a ``chilling effect'' on neighborhoods' willingness to reach out for help to local governments (Lerman and Weaver \protect\hyperlink{ref-Lerman2013}{2013}). Perhaps most troubling of all, these effects are not concentrated in random neighborhoods; as Gelman, Fagan, and Kiss (\protect\hyperlink{ref-Gelman2007}{2007}) shows, for instance, New York's ``stop-and-frisk'' policy impacted Black and Latino New Yorkers at rates far higher than Whites, even after controlling for neighborhood variability and race-specific arrest rates. The result is an incarcerated population that looks far different than the rest of th state: according to data from the New York State Department of Corrections and Community Supervision, 49.5 percent of individuals who were incarcerated in December of 2018 were non-Hispanic Black, although the Census Bureau estimates that just 14.3 percent of the citizen voting age population in the state is non-Hispanic Black.\footnote{Latinos are also over-represented among the incarcerated population, though not as dramatically: Latinos make up 14.1 percent of the citizen voting age population and 22.9 percent of the incarcerated population.}

Felony disenfranchisement policies are part of a criminal justice system that disproportionately impacts Black Americans living in certain communities. The effects of disenfranchisement are concentrated in neighborhoods that already suffer from myriad disadvantages thanks to social and economic marginalization. The neighborhood-specific implications of felony disenfranchisement, however, remain largely unstudied. A number of studies have explored the effect of imprisonment and disenfranchisement on later political participation (White \protect\hyperlink{ref-White2019}{2019}\protect\hyperlink{ref-White2019}{a}; Gerber et al. \protect\hyperlink{ref-Gerber2014}{2014}; Burch \protect\hyperlink{ref-Burch2011}{2011}). Others have looked at the spillover effects of disenfranchisement on eligible Black voters at the state level (Bowers and Preuhs \protect\hyperlink{ref-Bowers2009}{2009}; King and Erickson \protect\hyperlink{ref-King2016}{2016}). With the exception of Burch (\protect\hyperlink{ref-Burch2013}{2013}), however, little attention has been paid to the impact of felony disenfranchisement on political participation at the neighborhood level.

Filling this gap in the literature is of great importance when we consider the spatial concentration of policing networks. If the spillover effects of felony disenfranchisement are widely dispersed, the political consequences are likely similarly diffuse. If, however, the effects are both large enough to be detected at a statewide level (as the literature suggests) but also concentrated in a few neighborhoods, disenfranchisement laws likely severely undermine local political representation This lowered participation can disadvantage neighborhoods insofar as they have unique candidate preferences. It could also lower the resources distributed to these neighborhoods, as politicians determine that too few votes can be won through investing in these areas. This is not merely of theoretical concern: research indicates that turnout differentials at the city level have material consequences. As Hajnal (\protect\hyperlink{ref-Hajnal2009}{2009}) tells us: ``low {[}and uneven{]} turnout results in losses in mayoral elections, less equitable racial and ethnic representation on city councils, and spending policies that are less in line with the preferences of racial and ethnic minorities and other disadvantaged groups'' (8).

@@

Since the 2000 election, scholars have attempted to quantify the effect of felony disenfranchisement on the political representation of highly-incarcerated communities. Uggen and Manza (\protect\hyperlink{ref-Uggen2002}{2002}) produced the first estimates of felony disenfranchisement's impact on turnout, arguing that Al Gore would have won the presidency if not for for disenfranchisment in Florida. Their analysis estimated the direct effects of felony disenfranchisement on turnout by quantifying the number of actually disenfranchised individuals who would have participated if given the chance. Since Uggen and Manza (\protect\hyperlink{ref-Uggen2002}{2002}), other scholars have also investigated the direct effect of felony disenfranchisement (Miles \protect\hyperlink{ref-Miles2004}{2004}; Uggen and Manza \protect\hyperlink{ref-Uggen2004}{2004}; Drucker and Barreras \protect\hyperlink{ref-Drucker2005}{2005}; Ochs \protect\hyperlink{ref-Ochs2006}{2006}).

A number of papers have also explored the indirect impact felony disenfranchisement policies have on turnout among non-disenfranchised residents. King and Erickson (\protect\hyperlink{ref-King2016}{2016}), for instance, leverages state-level variation in disenfranchisement laws to estimate the impact that felony disenfranchisement has on turnout among Black Americans. They use data from the 2004 Current Population Survey Voting and Registration Supplement to calculate statewide turnout rates, and include estimates of the share of Black Americans who are disenfranchised in each state from Manza and Uggen (\protect\hyperlink{ref-locked_out}{2006}) to explore the impact of these policies on eligible voters. They conclude that disenfranchisement has large spillover effects for Black voters: where more Black residents are disenfranchised, eligible Black voters are less likely to cast a ballot. These findings are in line with other research that has explored whether the effects of disenfranchisement extend beyond those whose voting rights are directly suspended (Bowers and Preuhs \protect\hyperlink{ref-Bowers2009}{2009}; Ochs \protect\hyperlink{ref-Ochs2006}{2006}). As Bowers and Preuhs (\protect\hyperlink{ref-Bowers2009}{2009}) sums up: ``{[}I{]}t is not solely the direct vote of ex-felons that is denied through these laws. {[}Felony disenfranchisement{]} impacts the political power of communities that extends beyond felons' collateral penalty'' (724).

Although scholars have established that felony disenfranchisement decreases turnout among Black voters at the \emph{state} level, relatively little research has been done on how felony disenfranchisement operates at the sub-state level. Though we know that Black voters are generally less likely to cast a ballot when they live in a state with strict disenfranchisement laws, little work has been done exploring the impact these laws might have at the local level. Burch (\protect\hyperlink{ref-Burch2013}{2013}) is an exception to this. Burch explores the depressive effect of disenfranchisement laws at the local level in North Carolina by examining census block group level turnout and zip code level involvement with the criminal justice system, determining that ``at high concentrations, imprisonment and community supervision have an unequivocally demobilizing effect of neighborhoods'' (185). This project seeks to expand on her work by replicating her findings in New York City; by using a different estimation technique; and by using a different definition of ``lost voters.''

@@

Recent work from Hannah Walker and others, however, leads me to moderate my hypothesis about the indirect neighborhood effects of lost voters. Walker (\protect\hyperlink{ref-Walker2014}{2014}) and Walker and Garcı́a-Castañon (\protect\hyperlink{ref-Walker2017}{2017}), for instance, demonstrate that individuals who have proximal contact with the criminal justice system (defined as ``as having a loved one who is a custodial citizen without yourself having had contact'' (Walker and Garcı́a-Castañon \protect\hyperlink{ref-Walker2017}{2017}, 542)) were not less likely to vote, but \emph{were} more likely to participate in nonelectoral political events (such as signing a petition, attending a community meeting, or writing to an elected official). These effects are particularly pronounced for women of color. Although Walker and others demonstrate that these actions largely take place outside of the voting booth, they may moderate spillover effects on turnout.

Similarly, White (\protect\hyperlink{ref-White2019a}{2019}\protect\hyperlink{ref-White2019a}{b}) finds that incarceration has only moderate indirect effects on voting. She finds ``evidence of a short-term demobilization effect for people who see household members convicted or jailed in the weeks before the election, but no evidence of a lasting turnout effect from these experiences'' (607). This study does not, however, interrogate whether the incarceration of a would-be voter has different indirect effects than the incarceration of an individual who would not have voted either way. It is possible that the indirect effects of the incarceration or jailing of a would-be voter are different than the indirect effects arising from other individuals. Nevertheless, the literature on the effects of proximal contact with the criminal justice system and felony disenfranchisement is hardly settled.

\hypertarget{theoretical-implications-for-reduced-turnout}{%
\section*{Theoretical Implications for Reduced Turnout}\label{theoretical-implications-for-reduced-turnout}}
\addcontentsline{toc}{section}{Theoretical Implications for Reduced Turnout}

As discussed above, it has been widely established that felony disenfranchisement reduces turnout even among eligible voters, particularly in the Black community (e.g.~King and Erickson \protect\hyperlink{ref-King2016}{2016}; Bowers and Preuhs \protect\hyperlink{ref-Bowers2009}{2009}). With the exception of Burch (\protect\hyperlink{ref-Burch2013}{2013}), however, there has been little investigation into whether these depressive effects are narrowly concentrated in the neighborhoods home to disenfranchised individuals or are more widely dispersed. Understanding how these spillover effects are distributed is of practical importance. For instance, if neighborhoods home to disenfranchised individuals display different candidate preferences than the rest of the city, these demobilizing effects will undermine these neighborhoods' political representation. There is some evidence that spatial segregation leads to unique voting preferences among neighborhoods: Kinsella, McTague, and Raleigh (\protect\hyperlink{ref-Kinsella2015}{2015}) examines political clustering in the Greater Cincinnati Metropolitan Area from 1976 through 2008. Over this three-decade period, precinct-level presidential election results show trends toward increased polarization.

Even if neighborhoods do not have unique preferences for candidates, lower turnout in impacted communities may reduce their allocation of public goods. At the Congressional level representatives direct resources to areas within their districts that provide the greatest political benefit --- that is to say, areas that will reward them with more votes (Martin \protect\hyperlink{ref-Martin2003}{2003}). Congressional representatives are also more responsive to the policy preferences of higher-turnout areas. ``{[}H{]}igher citizen participation is rewarded,'' Martin and Claibourn (\protect\hyperlink{ref-Martin2013}{2013}) concludes, ``with enhanced policy responsiveness'' (59). Griffin and Newman (\protect\hyperlink{ref-Griffin2005}{2005}) finds similar effects in the United States Senate, demonstrating that ``voter preferences predict the aggregate roll-call behavior of Senators while nonvoter preferences do not'' (1206). If the neighborhoods most impacted by felony disenfranchisement turn out at lower rates, they may find that their elected representatives are less likely to support their needs and support their calls for greater public investment for amenities such as schools and parks. This may be true even if they support the same candidates as neighborhoods not touched by felony disenfranchisement.

Although research on the impact of local turnout on city-wide policy is scarce, Anzia (\protect\hyperlink{ref-Anzia2019}{2019}) examines the impact of senior turnout on ``senior-friendly'' policy at the city level. Anzia does not find that senior turnout in general increases the likelihood of senior-friendly policies, but that elected officials \emph{are} responsive to senior turnout when seniors ``are a more cohesive, meaningful group'' (1). American cities are highly segregated by race and by class, but less so by age. As such, Anzia's study does not speak directly to the impact of \emph{neighborhood} turnout rates on city policy.

Hajnal and Trounstine (\protect\hyperlink{ref-Hajnal2005}{2005}) and Hajnal (\protect\hyperlink{ref-Hajnal2009}{2009}), however, show that racial variation in turnout has real consequences for local elections and local political power. They estimate whether mayoral and city council races might have turned out differently if all races had turned out at the same rate, finding that three of the ten largest cities in America would likely have had different mayors if minorities had turned out as frequently as Whites. Of particular interest for the study at hand is their assertion that New York City would have elected a different mayor in 2001 if all racial groups had participated equally. Hajnal and Trounstine (\protect\hyperlink{ref-Hajnal2005}{2005}) tells us that ``Changes in the percentage of voters who turn out can and do alter mayoral election outcomes and racial representation on city councils'' (518). Neighborhoods most impacted by felony disenfranchisement are disproportionately home to residents of color, and they are likely to share concerns about criminal justice issue. Hajnal and Trounstine (\protect\hyperlink{ref-Hajnal2005}{2005}) and Anzia (\protect\hyperlink{ref-Anzia2019}{2019}) both therefore indicate that depressed turnout from felony disenfranchisement is likely to materially reduce neighborhoods' political representation.

\hypertarget{data}{%
\section*{Data}\label{data}}
\addcontentsline{toc}{section}{Data}

\hypertarget{criminal-justice-data}{%
\subsection*{Criminal Justice Data}\label{criminal-justice-data}}
\addcontentsline{toc}{subsection}{Criminal Justice Data}

The primary criminal justice dataset comes from a public records request filed with the New York State Department of Corrections and Community Supervision (NYSDOCCS). They include individual-level incarceration and parole records for individuals who have been incarcerated in New York State since 1990. The data includes a host of information, including: first, middle, and last name; date of birth; class of offense; incarceration start and end dates; dates of parole; sex; race; and others. These data are used to determine when individuals were incarcerated or on parole, the class of crimes for which they were incarcerated, and other demographic information.

The state makes records available only for individuals who have been incarcerated for felony offenses. It does not make information about individuals sentenced to probation or incarcerated for misdemeanors. Thus, while the data covers all individuals subject to felony disenfranchisement rules (only individuals incarcerated for felony offenses lose their voting rights), it limits the availability of a potentially helpful control group. It does not include individuals who are held in federal prisons; however, because the vast majority of incarcerated felons are held in state prisons, this is unlikely to affect the analysis.

\hypertarget{voter-file-data}{%
\subsection*{Voter File Data}\label{voter-file-data}}
\addcontentsline{toc}{subsection}{Voter File Data}

Most states in the United States are required to maintain files with information on all registered voters. In New York, this information is publicly available from the Board of Elections. It includes information on all registered voters, including: first, middle, and last name; date of birth; vote history; and other information. The New York State Voter File also includes information on voters who were previously registered but have since been purged, either because they moved, died, or were incarcerated for a felony offense. I use a snapshot of the registered voter file from April 30\textsuperscript{th}, 2018.

The New York State Voter File is unique in its treatment of ``purged'' voters: although most states remove voters from their voter files once they are no longer eligible to vote, New York continues to include them in the file (but marks them as purged). I can therefore identify voters who were registered in the past but have since been purged due to a felony conviction.

\hypertarget{geocoding}{%
\subsection*{Geocoding}\label{geocoding}}
\addcontentsline{toc}{subsection}{Geocoding}

Voters' home addresses were converted to latitudes and longitudes using a geocoder provided by SmartyStreets. I then used the statistical software R to map these latitudes and longitudes to census block groups, census tracts, and city council districts using shapefiles publicly available from the Census Bureau and the City of New York. This geocoder is not perfect: among individuals registered to vote in New York City, the geocoder failed to determine the latitude and longitude of the addresses of 1 percent of registered voters. The geocoder was slightly less successful when it came to lost voters (defined and discussed below); 1.6 percent of these individuals were not geocoded. Voters who were not successfully geocoded are dropped from the dataset; however, because so few observations went uncoded, it is unlikely to affect the analysis.

\hypertarget{matching}{%
\subsection*{Matching}\label{matching}}
\addcontentsline{toc}{subsection}{Matching}

Registered and formerly registered individuals who have been to prison in New York are identified by matching the NYSDOCCS records with the registered voter file. I match individuals in each dataset using first name, middle name, last name, and date of birth. To be considered a ``match,'' records must have the exact same birth date. The first and last names must also be exact matches (conditional on the adjustments discussed below). The middle names must meet one of the following conditions in order to qualify:

\begin{itemize}
\tightlist
\item
  Middle names are identical. If neither set of records includes a middle name, this condition is met.
\item
  A full middle name in one set of records and only a middle initial in the other. The first letter of the full middle name must be the same as the middle initial in the other set of records.
\item
  A middle name or middle initial in one set of records, and a missing middle name in the other set.
\end{itemize}

Thus, ``John Andrew Doe'' and ``John A Doe'' would count as matches. Similarly, ``John Andrew Doe'' and ``John Doe'' would count, while ``John Andrew Doe'' and ``John Anthony Doe'' would not.

There are two types of potential error in this methodology: a false positive will result when a formerly incarcerated individual's records matches the record of a voter who is a different individual but shares the same name and date of birth. False negatives will occur when an individual has a different name in the different sets of records, or when the birthdate is incorrectly reported in one of the sets of records

Testing for the presence of false positive matches is fairly straightforward. Meredith and Morse (\protect\hyperlink{ref-Meredith2013}{2013}) offers one way to test their prevalence using placebo matching. I slightly alter the date of birth reported in the NYSDOCCS dataset to create false records. Comparing the number of matches between these ``fake'' records and the voter file with the number of matches between the ``true'' records and the voter file provides an estimate of how frequently false positives occur. Table \ref{tab:change-dobs} shows the results of true matches, as well as matches using a set of fake records created by adding or subtracting 35 days from an individual's birthdate. This analysis indicates that false positives account for between 0.4 and 0.5 percent of all matches, a share that is likely too small to have any material impact on the overall analysis. The numbers in Table \ref{tab:change-dobs} are derived by matching (and modifying) all individuals who were incarcerated or on parole on Election Day in 2017 with the registered voter file from April of 2018.

\begin{table}[H]

\caption{\label{tab:shift-dobs-chunk}\label{tab:change-dobs} Results of Shifting Birthdates}
\centering
\begin{tabular}[t]{cc}
\toprule
Group & \makecell[l]{Number of Matches Between\\DOCCS and Voter File Records}\\
\midrule
Actual Birthdate & 20,955\\
Birthdate + 35 Days & 105\\
Birthdate - 35 Days & 92\\
\bottomrule
\end{tabular}
\end{table}

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
  12pt,
]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Neighborhoods and Felony Disenfranchisement: The Case of New York City},
  pdfauthor={Kevin Morris},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{longtable,booktabs}
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\usepackage{rotating}
\newcommand{\beginsupplement}{\setcounter{table}{0}  \renewcommand{\thetable}{A\arabic{table}} \setcounter{figure}{0} \renewcommand{\thefigure}{A\arabic{figure}}}
\usepackage{setspace}\doublespacing
\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newenvironment{cslreferences}%
  {\setlength{\parindent}{0pt}%
  \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces}%
  {\par}

\title{Neighborhoods and Felony Disenfranchisement: The Case of New York City}
\author{Kevin Morris\footnote{The author thanks Jacob Faber, Jeff Manza, Myrna Pérez, Ariel White, and Peter Miller for their comments on this project. All errors are my responsibility.}}
\date{March 03, 2020}

@@

The political history of the United States has been characterized by a general, if nonlinear, trend toward universal suffrage (see, for instance, Keyssar \protect\hyperlink{ref-Keyssar2009}{2009}). At the time of the nation's founding, access to the ballot box was restricted to landed White men; over the following two centuries, the franchise was greatly expanded. Today, voting rights are considered foundational aspects of full citizenship (United Nations General Assembly Resolution 2200 (XXI)). Despite the United State's march toward ever-more-inclusive systems of democracy, however, one large group of American citizens is formally barred from voting. In most of the United States, citizens convicted of felonies are at least temporarily prohibited from casting ballots in elections (Brennan Center for Justice \protect\hyperlink{ref-bcj_laws}{2019}). Although some states such as Florida and Louisiana have gradually moved to dismantle their systems of felony disenfranchisement, an estimated 4.7 million American citizens remain disenfranchised (Uggen, Larson, and Shannon \protect\hyperlink{ref-sentencing_2016}{2016}).

Due to economic and racial segregation, these effects are highly spatially concentrated. Data available from New York City shows that in 2017, 10 of the New York Police Department's 77 precincts were responsible for more than a quarter of all arrests for felony charges. Many scholars have detailed the impact of living in areas with high levels of police activity. Residents of such neighborhoods suffer from worse physical health (Sewell and Jefferson \protect\hyperlink{ref-Sewell2016}{2016}) and are more likely to suffer from anxiety and exhibit symptoms of trauma (Geller et al. \protect\hyperlink{ref-Geller2014}{2014}). The labor markets and social networks in neighborhoods with high levels of policing and incarceration are disrupted (Clear \protect\hyperlink{ref-Clear2008}{2008}), while concentrated policing has also been credited with having a ``chilling effect'' on neighborhoods' willingness to reach out for help to local governments (Lerman and Weaver \protect\hyperlink{ref-Lerman2013}{2013}). Perhaps most troubling of all, these effects are not concentrated in random neighborhoods; as Gelman, Fagan, and Kiss (\protect\hyperlink{ref-Gelman2007}{2007}) shows, for instance, New York's ``stop-and-frisk'' policy impacted Black and Latino New Yorkers at rates far higher than Whites, even after controlling for neighborhood variability and race-specific arrest rates. The result is an incarcerated population that looks far different than the rest of the state: according to data from the New York State Department of Corrections and Community Supervision, 49.5 percent of individuals who were incarcerated in December of 2018 were non-Hispanic Black, although the Census Bureau estimates that just 14.3 percent of the citizen voting age population in the state is non-Hispanic Black.\footnote{Latinos are also over-represented among the incarcerated population, though not as dramatically: Latinos make up 14.1 percent of the citizen voting age population and 22.9 percent of the incarcerated population.}

Felony disenfranchisement policies are part of a criminal justice system that disproportionately impacts Black Americans living in certain communities. The effects of disenfranchisement are concentrated in neighborhoods that already suffer from myriad disadvantages thanks to social and economic marginalization. The neighborhood-specific implications of felony disenfranchisement, however, remain largely unstudied. A number of studies have explored the effect of imprisonment and disenfranchisement on later political participation (White \protect\hyperlink{ref-White2019}{2019}\protect\hyperlink{ref-White2019}{b}; Gerber et al. \protect\hyperlink{ref-Gerber2015}{2015}; Burch \protect\hyperlink{ref-Burch2011}{2011}). Others have looked at the spillover effects of disenfranchisement on eligible Black voters at the state level (Bowers and Preuhs \protect\hyperlink{ref-Bowers2009}{2009}; King and Erickson \protect\hyperlink{ref-King2016}{2016}). With the exception of Burch (\protect\hyperlink{ref-Burch2013}{2013}), however, little attention has been paid to the impact of felony disenfranchisement on political participation at the neighborhood level.

@@

Since the 2000 election, scholars have attempted to quantify the effect of felony disenfranchisement on the political representation of highly-incarcerated communities. Uggen and Manza (\protect\hyperlink{ref-Uggen2002}{2002}) produced the first estimates of felony disenfranchisement's impact on turnout, arguing that Al Gore would have won the presidency if not for disenfranchisement in Florida. Their analysis estimated the direct effects of felony disenfranchisement on turnout by quantifying the number of actually disenfranchised individuals who would have participated if given the chance. Since Uggen and Manza (\protect\hyperlink{ref-Uggen2002}{2002}), other scholars have also investigated the direct effect of felony disenfranchisement (Miles \protect\hyperlink{ref-Miles2004}{2004}; Uggen and Manza \protect\hyperlink{ref-Uggen2004}{2004}; Drucker and Barreras \protect\hyperlink{ref-Drucker2005}{2005}; Ochs \protect\hyperlink{ref-Ochs2006}{2006}).

@@

Recent work from Hannah Walker and others, however, leads me to moderate my hypothesis about the indirect neighborhood effects of lost voters. Walker (\protect\hyperlink{ref-Walker2014}{2014}) and Walker and García-Castañon (\protect\hyperlink{ref-Walker2017}{2017}), for instance, demonstrate that individuals who have proximal contact with the criminal justice system (defined as ``as having a loved one who is a custodial citizen without yourself having had contact'' (Walker and García-Castañon \protect\hyperlink{ref-Walker2017}{2017}, 542)) were not less likely to vote, but \emph{were} more likely to participate in nonelectoral political events (such as signing a petition, attending a community meeting, or writing to an elected official). These effects are particularly pronounced for women of color. Although Walker and others demonstrate that these actions largely take place outside of the voting booth, they may moderate spillover effects on turnout.

Similarly, White (\protect\hyperlink{ref-White2019a}{2019}\protect\hyperlink{ref-White2019a}{a}) finds that incarceration has only moderate indirect effects on voting. She finds ``evidence of a short-term demobilization effect for people who see household members convicted or jailed in the weeks before the election, but no evidence of a lasting turnout effect from these experiences'' (607). This study does not, however, interrogate whether the incarceration of a would-be voter has different indirect effects than the incarceration of an individual who would not have voted either way. It is possible that the indirect effects of the incarceration or jailing of a would-be voter are different than the indirect effects arising from other individuals. Nevertheless, the literature on the effects of proximal contact with the criminal justice system and felony disenfranchisement is hardly settled.

\hypertarget{methodology-and-data}{%
\section*{Methodology and Data}\label{methodology-and-data}}
\addcontentsline{toc}{section}{Methodology and Data}

To identify lost voters, and to determine whether lost voters reduce neighborhood turnout, I use data provided directly by the State of New York.

The first dataset is from the Department of Corrections and Community Supervision (DOCCS). This data includes information on every individual who was incarcerated in state prison, or on parole after serving time in state prison. These data include individuals' first, middle, and last names, as well as their date of birth, the dates they spent incarcerated or on parole, and other information. I limit the records in the data to include all individuals who were either in prison or on parole as of the mayoral election in 2017.

The second primary dataset is the registered voter file from New York State. The registered voter file includes information on every voter registered in the state such as name, address, vote history, and gender. It also includes the records of everyone who has been purged --- in other words, whose registration has been cancelled. The records are marked as ``purged,'' and allow me to identify the vote history even of individuals who have been purged for a felony conviction.

By matching these two datasets using first, middle, and last names, and date of birth, I can identify all individuals who were disenfranchised in the City of New York on election day in 2017 \emph{who had a history of voting.} Specifically, a ``lost voter'' is someone who cast a ballot between January 1, 2007 and December 31, 2016 and was imprisoned or on parole for a felony offense on November 7\textsuperscript{th}, 2017.

\hypertarget{why-do-we-care-about-lost-voters}{%
\section*{Why Do We Care About ``Lost Voters''?}\label{why-do-we-care-about-lost-voters}}
\addcontentsline{toc}{section}{Why Do We Care About ``Lost Voters''?}

In this analysis, I offer a different definition of ``lost voter'' than much of the literature. Many recent papers have attempted to identify relationships between the number of disenfranchised residents --- \emph{potentially} lost voters --- and turnout. Such an approach is informative for understanding the impact of disenfranchisement. Many young men, for instance, are admitted to prison each year. If someone is incarcerated shortly after they turn 18 in an odd-numbered year, they may be incarcerated before they even have the opportunity to cast a ballot. Similarly, we know that older individuals are more likely to vote; if someone who is incarcerated would have ``aged in'' to voting but for their incarceration, they will have no history of voting. This analytical approach is also often taken due to data constraints: most states' registered voter files provide only a snapshot of currently registered voters, making it impossible to determine voting history for individuals who are currently disenfranchised (and therefore not currently registered). Similarly, state-level data does not report the number of incarcerated individuals with a history of voting; therefore, studies that leverage variation in laws between states to estimate the effect of felony disenfranchisement rely on estimates of the total disenfranchised population, not the number of disenfranchised individuals with a history of voting.
